Nathan Greenslit Data Analysis Project

### 2023-04-05 products/manuscript/Manuscript.qmd
work cited

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year [@dalmeida1986]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico (Fig. 1). These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria [@kellogg2004], fungi [@ramirez2022], virus-like particles ([Griffin et al., 2001](https://link.springer.com/article/10.1023/A:1011868218901)), and minerals ([Formenti et al., 2003](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2002JD002648)). These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron ([Graham & Duce, 1967](https://www.sciencedirect.com/science/article/pii/0004698182901986), [Mills et al., 2004](https://pubmed.ncbi.nlm.nih.gov/15152251/), [Savoie et al., 1989](https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/JD094iD04p05069?casa_token=PMRaet9QSYsAAAAA:nb8yBorxdKaEWwGfI5aNILJ0D4h6drNu2YY1y1y8ExbQ03z59EUTQTOHoxinzsxRtR4jzsfSA73_Hig)). This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups ([Lenes et al., 2008](https://www.sciencedirect.com/science/article/pii/S0278434308000575), [Westrich et al., 2016](https://www.pnas.org/doi/abs/10.1073/pnas.1518080113), [Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)).

Previous work conducted in the West Florida Shelf has shown that increased iron availability can stimulate the growth of the nitrogen-fixing cyanobacterium *Trichodesmium*. The high amount of nitrogen produced via fixation can be adequate enough to stimulate toxic red tides, caused by the dinoflagellate *Karenia brevis* ([Lenes et al., 2008](https://www.sciencedirect.com/science/article/pii/S0278434308000575)).

While it is documented that these dust events can stimulate growth of harmful phytoplankton, emerging evidence shows that heterotrophic bacteria are responding to the episodic influx of limiting resources and substrate, resulting in blooms of bacteria that are associated with disease, which we refer to as harmful bacterial blooms. Among those that respond are bacterium belonging to the genus *Vibrio*, a group of ubiquitous marine opportunistic heterotrophs. While considered conditionally rare (comprising of \<1% of the microbial community), under the proper conditions, this genus can rapidly bloom to make up a large percentage of the community over a short period of time ([Westrich et al., 2016](https://www.pnas.org/doi/abs/10.1073/pnas.1518080113), [Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)). *Vibrio* species are found in brackish to marine waters with salinities ranging from 0-35 and will grow in waters with temperatures ranging from 5°C-40°C ([Sampaio et al., 2022](https://www.mdpi.com/1424-2818/14/2/97)). Each year, plumes of Saharan dust travel across the Atlantic via easterly trade winds. These plumes are then deposited in the surface waters of the mid-Atlantic, Caribbean, and Gulf of Mexico. These dust aerosols harbor a wide range of fungi, bacteria, virus-like-particles, minerals, and nutrients (NO3, Fe, PO4). Addition of these nutrients to otherwise oligotrophic settings can result in large and rapid blooms of potentially harmful microbes, presenting a danger to both marine and human health.

*Vibrio* species are excellent opportunists, having multiple advantages that allow them to quickly respond to any newly introduced substrate or nutrients. These advantages include multiple copies of rRNA genes, a rapid doubling time, efficient chemotactic motility, large genomic repertoire (consisting of two circular chromosomes), and multiple Fe-siderophore complexes that allow for rapid uptake of iron into the cell ([Jensen et al., 2009](https://academic.oup.com/femsle/article/294/2/207/512983), [Ringgaard et al., 2018](https://journals.asm.org/doi/10.1128/JB.00793-17), [Eagon, 1962](https://journals.asm.org/doi/abs/10.1128/jb.83.4.736-737.1962?casa_token=WdtRNnQWyhkAAAAA:Zre66xvwgEu3wXQykenhnSoy3OA08h1u-4Zl31k9cxxxfV6jq46dzL18Y02OmwPHdY7p7tgH0wVN-4I), [Okada et al., 2005](https://journals.asm.org/doi/10.1128/JB.187.2.752-757.2005), [Payne et al., 2015](https://journals.asm.org/doi/full/10.1128/MMBR.00046-15)).

Within the *Vibrionaceae* family, there are at least 12 species that are commonly known as human pathogens. Illness can be induced through either a foodborne route via the consumption of raw or uncooked seafood and contaminated drinking water, or through nonfoodborne routes such as wound exposure while swimming. The most common illnesses caused by *Vibrio* include self-limiting diarrhea and cholera (*V. cholerae*), shellfish-induced gastroenteritis (*V. parahaemolyticus, V. vulnificus*), extreme cases of necrotizing fasciitis and septicemia (*V. vulnificus*), and wound and ear infections  (*V. alginolyticus*, *V. vulnificus*) [@baker2018]. Some species also pose a threat to marine health, inducing mortality in oyster and clam larvae (*V. tubiashii, V. coralliilyticus*), bacterial bleaching and rapid tissue loss in corals (*V. mediterranei, V. coralliilyticus*) ([Richards et al., 2015](https://pubmed.ncbi.nlm.nih.gov/25344234/), [Rosenberg & Falkovitz, 2004](https://www.annualreviews.org/doi/10.1146/annurev.micro.58.030603.123610), [@haim2003]), and mortality in important aquaculture species such as penaeid shrimp (*V. parahaemolyticus*) and sea breams (*V. harveyi)* ([Tran et al., 2013](https://www.int-res.com/abstracts/dao/v105/n1/p45-55/), [Haldar et al., 2010](https://www.sciencedirect.com/science/article/pii/S0944501309001177)), among others.

Previous work has characterized *Vibrio* blooms in response to Saharan dust input in the oligotrophic setting of the Caribbean and subtropical Atlantic (Barbados and Florida Keys, respectively), with surface water concentrations of total *Vibrio* increasing by five to thirty times that found during non-dust conditions and returning to baseline levels within 24-48 hours (Fig. 2). *Vibrio* composition within the larger microbial community also shifted following dust deposition, with initial levels of \<1.4% to a peak of 19.8% of the bacterial community ([Westrich et al., 2016](https://www.pnas.org/doi/abs/10.1073/pnas.1518080113)). A similar phenomenon was also observed in the surface waters of the tropical and open ocean mid-Atlantic with *Vibrio* populations increasing 1.5-fold in the mid-Atlantic following deposition ([Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)). In a follow up study in the Florida Keys, episodic dust events during a daily time series promoted a succession of bacterial responses, with declines in *Prochlorococcus* coinciding with initial increases in bacteria belonging to the order *Vibrionales*, followed by subsequent shifts in response of different bacterial groups ([Borchardt et al., 2019](https://aslopubs.onlinelibrary.wiley.com/doi/full/10.1002/lno.11291)). To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients. However, the composition of desert dust can be complex and may have the potential to deliver critical resources that could be exploited by opportunistic microbes (like *Vibrio*) and elicit a growth response despite higher baseline nutrient levels. Given that human exposure is likely in these coastal waters, it is important to elucidate the relationship between dust input and microbial blooms in these higher nutrient environments.

@@

Sampling took place at three locations in Corpus Christi, TX: Blind Oso Bay, a residential canal system on Padre Island, and the Gulf. Sites were chosen based on differing baseline nutrient levels. Blind Oso Bay is a shallow tributary that is popular amongst wadefisherman and kayakers. The Bay is often impacted by fluctuating salinities and high levels of nutrients and chlorophyl due to its proximity to a wastewater treatment plant ([Wetz, 2014](https://www.cbbep.org/manager/wp-content/uploads/1312-Oso-Bay-Nutrient-Assessment-Final_Report-6-24-14.pdf)). Additionally, the Bay has persistent issues with high levels of fecal indicator bacteria, placing it on the U.S. impaired waters list ([Nicolau & Hill, 2014](https://www.tceq.texas.gov/downloads/water-quality/tmdl/oso-bay-oyster-harvesting-assessment-103/103-oso-bay-monitoring-report-2013.pdf), [Texas Commission on Environmental Quality, 2022](https://www.tceq.texas.gov/downloads/water-quality/assessment/integrated-report-2022/2022-303d.pdf)). The residential canal system is also impacted by high nutrients and chlorophyl from storm water runoff, and salinity is primarily driven by precipitation. The Gulf site is characterized by constant salinities (\~36) and low nutrient levels.

@@

Total *Vibrio* were estimated using genus-specific quantitative polymerase chain reaction (qPCR) ([Westrich et al., 2018](https://www.frontiersin.org/articles/10.3389/fmars.2018.00012/full)). <!--# Add specific primers, conditions, etc  -->

>>>>

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year [@dalmeidaModelSaharanDust1986]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico (Fig. 1). These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria [@kelloggCharacterizationAerosolizedBacteria2004], fungi [@ramirez-camejoFungalDiversitySahara2022], virus-like particles [@griffinAfricanDesertDust], and minerals [@formenti2003]. These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron [@millsIronPhosphorusColimit2004 ; @savoieNonseasaltSulfateNitrate1989 ; @graham1982] . This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups [@lenesSaharanDustPhosphatic2008; @westrichSaharanDustNutrients2016a ; @westrichSaharanDustNutrients2016a].

Previous work conducted in the West Florida Shelf has shown that increased iron availability can stimulate the growth of the nitrogen-fixing cyanobacterium *Trichodesmium*. The high amount of nitrogen produced via fixation can be adequate enough to stimulate toxic red tides, caused by the dinoflagellate *Karenia brevis* [@lenesSaharanDustPhosphatic2008].

While it is documented that these dust events can stimulate growth of harmful phytoplankton, emerging evidence shows that heterotrophic bacteria are responding to the episodic influx of limiting resources and substrate, resulting in blooms of bacteria that are associated with disease, which we refer to as harmful bacterial blooms. Among those that respond are bacterium belonging to the genus *Vibrio*, a group of ubiquitous marine opportunistic heterotrophs. While considered conditionally rare (comprising of \<1% of the microbial community), under the proper conditions, this genus can rapidly bloom to make up a large percentage of the community over a short period of time [@westrichSaharanDustNutrients2016a ; @westrichVibrioPopulationDynamics2018a]. *Vibrio* species are found in brackish to marine waters with salinities ranging from 0-35 and will grow in waters with temperatures ranging from 5°C-40°C [@sampaioVibrioSppLife2022]. Each year, plumes of Saharan dust travel across the Atlantic via easterly trade winds. These plumes are then deposited in the surface waters of the mid-Atlantic, Caribbean, and Gulf of Mexico. These dust aerosols harbor a wide range of fungi, bacteria, virus-like-particles, minerals, and nutrients (NO3, Fe, PO4). Addition of these nutrients to otherwise oligotrophic settings can result in large and rapid blooms of potentially harmful microbes, presenting a danger to both marine and human health.

*Vibrio* species are excellent opportunists, having multiple advantages that allow them to quickly respond to any newly introduced substrate or nutrients. These advantages include multiple copies of rRNA genes, a rapid doubling time, efficient chemotactic motility, large genomic repertoire (consisting of two circular chromosomes), and multiple Fe-siderophore complexes that allow for rapid uptake of iron into the cell [@jensenNonrandomMicroheterogeneity16S2009; @ringgaardChemotaxisArraysVibrio2018 ; @eagonPSEUDOMONASNATRIEGENSMARINE1962 ; @okadaVibriosCommonlyPossess2005 ; @payneVibrioIronTransport2016].

Within the *Vibrionaceae* family, there are at least 12 species that are commonly known as human pathogens. Illness can be induced through either a foodborne route via the consumption of raw or uncooked seafood and contaminated drinking water, or through nonfoodborne routes such as wound exposure while swimming. The most common illnesses caused by *Vibrio* include self-limiting diarrhea and cholera (*V. cholerae*), shellfish-induced gastroenteritis (*V. parahaemolyticus, V. vulnificus*), extreme cases of necrotizing fasciitis and septicemia (*V. vulnificus*), and wound and ear infections  (*V. alginolyticus*, *V. vulnificus*) [@baker-austinVibrioSppInfections2018]. Some species also pose a threat to marine health, inducing mortality in oyster and clam larvae (*V. tubiashii, V. coralliilyticus*), bacterial bleaching and rapid tissue loss in corals (*V. mediterranei, V. coralliilyticus*) [@richardsMortalitiesEasternPacific2015 ; @rosenbergVibrioShiloiOculina2004; @ben-haimVibrioCoralliilyticusSp2003], and mortality in important aquaculture species such as penaeid shrimp (*V. parahaemolyticus*) and sea breams (*V. harveyi)* [@tranDeterminationInfectiousNature2013 ; @haldarIdentificationVibrioHarveyi2010] among others.

Previous work has characterized *Vibrio* blooms in response to Saharan dust input in the oligotrophic setting of the Caribbean and subtropical Atlantic (Barbados and Florida Keys, respectively), with surface water concentrations of total *Vibrio* increasing by five to thirty times that found during non-dust conditions and returning to baseline levels within 24-48 hours (Fig. 2). *Vibrio* composition within the larger microbial community also shifted following dust deposition, with initial levels of \<1.4% to a peak of 19.8% of the bacterial community [@westrichSaharanDustNutrients2016a]. A similar phenomenon was also observed in the surface waters of the tropical and open ocean mid-Atlantic with *Vibrio* populations increasing 1.5-fold in the mid-Atlantic following deposition [@westrichVibrioPopulationDynamics2018a] In a follow up study in the Florida Keys, episodic dust events during a daily time series promoted a succession of bacterial responses, with declines in *Prochlorococcus* coinciding with initial increases in bacteria belonging to the order *Vibrionales*, followed by subsequent shifts in response of different bacterial groups [@borchardtSaharanDustDeposition2020b]. To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients. However, the composition of desert dust can be complex and may have the potential to deliver critical resources that could be exploited by opportunistic microbes (like *Vibrio*) and elicit a growth response despite higher baseline nutrient levels. Given that human exposure is likely in these coastal waters, it is important to elucidate the relationship between dust input and microbial blooms in these higher nutrient environments.

@@

Sampling took place at three locations in Corpus Christi, TX: Blind Oso Bay, a residential canal system on Padre Island, and the Gulf. Sites were chosen based on differing baseline nutrient levels. Blind Oso Bay is a shallow tributary that is popular amongst wadefisherman and kayakers. The Bay is often impacted by fluctuating salinities and high levels of nutrients and chlorophyl due to its proximity to a wastewater treatment plant [@wetz] Additionally, the Bay has persistent issues with high levels of fecal indicator bacteria, placing it on the U.S. impaired waters list ([Nicolau & Hill, 2014](https://www.tceq.texas.gov/downloads/water-quality/tmdl/oso-bay-oyster-harvesting-assessment-103/103-oso-bay-monitoring-report-2013.pdf), [Texas Commission on Environmental Quality, 2022](https://www.tceq.texas.gov/downloads/water-quality/assessment/integrated-report-2022/2022-303d.pdf)). The residential canal system is also impacted by high nutrients and chlorophyl from storm water runoff, and salinity is primarily driven by precipitation. The Gulf site is characterized by constant salinities (\~36) and low nutrient levels.

@@

Total *Vibrio* were estimated using genus-specific quantitative polymerase chain reaction (qPCR) [@westrichSaharanDustNutrients2016a]. <!--# Add specific primers, conditions, etc  -->
